---
title: "Data Preprocessing"
author: "Josh Schechter"
date: "2025-04-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
```

# Import student dataset
```{r}
students <- read.csv("../data/Students_Grading_Dataset.csv")
```

# Remove rows with missing values
```{r}
students <- students[complete.cases(students), ]
students <- na.omit(students)
students <- students[students$Parent_Education_Level != "", ]
```

```{r}
students <- students %>%
  rename(
    Stress_Level = `Stress_Level..1.10.`,
    Attendance = `Attendance....`
  )
```

```{r}
students$Pass_Fail <- ifelse(students$Grade %in% c("A", "B", "C"), 1, 0)

students$Pass_Fail <- factor(students$Pass_Fail, levels = c(1, 0))

#table(students$Pass_Fail)
```

```{r}
# output to csv
write.csv(students, "../data/Students_Grading_Dataset_Cleaned.csv", row.names = FALSE)
```

```{r}
colnames(students)
nrow(students)
```

# Display the first few rows of the dataset
```{r}
head(students, 5)
```
# Features of interest:
## Student_ID, Grade, Study_Hours_per_Week, Family_Income_Level, Sleep_Hours_per_Night,
## Stress_Level, Attendance, Internet_Access_at_Home, Parent_Education_Level, Extracurricular_Activities
## Gender

# Summary statistics
```{r}
summary(select(students, Student_ID, Total_Score, Grade, Study_Hours_per_Week, Sleep_Hours_per_Night,
               Family_Income_Level, Study_Hours_per_Week, Stress_Level, Attendance, Internet_Access_at_Home,
               Parent_Education_Level, Extracurricular_Activities))
```



# Data Visualization

## Numerical
```{r}
# Grade Distribution
students %>%
  count(Grade) %>%
  mutate(Proportion = n / sum(n)) %>%
  ggplot(aes(x = Grade, y = Proportion)) +
  geom_bar(stat = "identity", fill = 'lightblue') +
  labs(title = "Grade Distribution", y = "Proportion", x = "Grade") +
  ylim(0, 0.5)

# Total Score Distribution
ggplot(students, aes(x = Total_Score)) +
  geom_histogram(fill = "lightblue", bins=20) +
  labs(title = "Total Score Distribution",
       x = "Total Score")

# Final Score Boxplot
ggplot(students, aes(y = Final_Score)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Final Score by Grade")

# Attendance by Grade
ggplot(students, aes(x = Grade, y = Attendance)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Attendance by Grade")

# Sleep Hours by Grade
ggplot(students, aes(x = Grade, y = Sleep_Hours_per_Night)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Sleep Hours by Grade")

# Stress by Grade
ggplot(students, aes(x = Grade, y = Stress_Level)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Stress Level by Grade")

# Study Time by Grade
ggplot(students, aes(x = Grade, y = Study_Hours_per_Week)) +
  geom_boxplot(fill = "lightblue") +
  labs(title = "Study Hours per Week by Grade")


```

## Categorical
```{r}
# Family Income by Grade
students %>%
  count(Family_Income_Level, Grade) %>%
  group_by(Family_Income_Level) %>%
  mutate(Proportion = n / sum(n)) %>%
  ggplot(aes(x = Family_Income_Level, y = Proportion, fill = Grade)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Normalized Grade Distribution by Family Income Level",
       x = "Family Income Level", y = "Proportion") +
  ylim(0,0.5)

# Parent Education Levels by Grade
students %>%
  count(Parent_Education_Level, Grade) %>%
  group_by(Parent_Education_Level) %>%
  mutate(Proportion = n / sum(n)) %>%
  ggplot(aes(x = Parent_Education_Level, y = Proportion, fill = Grade)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Normalized Grade Distribution by Parent Education Level",
       x = "Parent Education Level", y = "Proportion") +
  ylim(0,0.5)

# Internet Access by Grade
students %>%
  count(Internet_Access_at_Home, Grade) %>%
  group_by(Internet_Access_at_Home) %>%
  mutate(Proportion = n / sum(n)) %>%
  ggplot(aes(x = Internet_Access_at_Home, y = Proportion, fill = Grade)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Normalized Grade Distribution by Internet Access at Home",
       x = "Internet Access at Home", y = "Proportion") +
  ylim(0,0.5)

# Extracurricular by Grade
students %>%
  count(Extracurricular_Activities, Grade) %>%
  group_by(Extracurricular_Activities) %>%
  mutate(Proportion = n / sum(n)) %>%
  ggplot(aes(x = Extracurricular_Activities, y = Proportion, fill = Grade)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Normalized Grade Distribution by Extracurricular Participation",
       x = "Extracurricular Activities", y = "Proportion") +
  ylim(0,0.5)
```



Table of proportion of students by grade for each family income level
```{r}
# Proportion table
grade_income_table <- students %>%
  count(Family_Income_Level, Grade) %>%
  group_by(Family_Income_Level) %>%
  mutate(Proportion = n / sum(n)) %>%
  select(-n) %>%
  pivot_wider(names_from = Grade, values_from = Proportion, values_fill = 0)

grade_parentedu_table <- students %>%
  count(Parent_Education_Level, Grade) %>%
  group_by(Parent_Education_Level) %>%
  mutate(Proportion = n / sum(n)) %>%
  select(-n) %>%
  pivot_wider(names_from = Grade, values_from = Proportion, values_fill = 0)

grade_internet_table <- students %>%
  count(Internet_Access_at_Home, Grade) %>%
  group_by(Internet_Access_at_Home) %>%
  mutate(Proportion = n / sum(n)) %>%
  select(-n) %>%
  pivot_wider(names_from = Grade, values_from = Proportion, values_fill = 0)

grade_extracurricular_table <- students %>%
  count(Extracurricular_Activities, Grade) %>%
  group_by(Extracurricular_Activities) %>%
  mutate(Proportion = n / sum(n)) %>%
  select(-n) %>%
  pivot_wider(names_from = Grade, values_from = Proportion, values_fill = 0)

grade_income_table
grade_parentedu_table
grade_internet_table
grade_extracurricular_table
```


```{r}
qqnorm(students$Total_Score, main="Normal Q-Q Plot of Total Score")
qqline(students$Total_Score, col = "red")
```


## EDA Results:

- Class attendance seems to be the best indicator for success
- Students with family/home life backgrounds don't seem to have a significant difference in grade outcomes.


```{r}
roc(students$Pass_Fail, students$Total_Score)
```


## Most important figures are below
```{r}
# Boxplot: Final Grade by Parental Education
ggplot(students, aes(x = Parent_Education_Level, y = Total_Score)) +
  geom_boxplot() +
  labs(title = "Total Score vs Parental Education Level", x = "Parental Education Level", y = "Total Score") +
  theme_minimal()

# Boxplot: Final Grade by Family Income
ggplot(students, aes(x = Family_Income_Level, y = Total_Score)) +
  geom_boxplot() +
  labs(title = "Total Score vs Family Income", x = "Family Income", y = "Total Score") +
  theme_minimal()

# Boxplot: Final Grade by Internet Access
ggplot(students, aes(x = Internet_Access_at_Home, y = Total_Score)) +
  geom_boxplot() +
  labs(title = "Total Score vs Internet Access", x = "Internet Access (Yes/No)", y = "Total Score") +
  theme_minimal()

# Boxplot: Final Grade by Extracurricular Activities
ggplot(students, aes(x = Extracurricular_Activities, y = Total_Score)) +
  geom_boxplot() +
  labs(title = "Total Score vs Extracurricular Participation", x = "Extracurricular Activities (Yes/No)", y = "Total Score") +
  theme_minimal()

ggplot(students, aes(x = factor(Stress_Level), y = Total_Score)) +
  geom_boxplot() +
  labs(title = "Final Grade vs Stress Level",
       x = "Stress Level (1 = low stress, 10 = high stress)",
       y = "Total Score") +
  theme_minimal()

# Scatterplot: Final Grade vs Study Time
ggplot(students, aes(x = Study_Hours_per_Week, y = Total_Score)) +
  geom_point(alpha = 0.5) +
  labs(title = "Total Score vs Study Time", x = "Study Time (hours)", y = "Total Score") +
  theme_minimal()
```
