---
title: "Modeling"
author: "Josh Schechter"
date: "2025-04-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
library(caret)
```

```{r}
students <- read.csv("../data/Students_Grading_Dataset_Cleaned.csv")
```

```{r}
colnames(students)
```
## Steps:
1. Find features of interest of interest.
2. Create individual regression models for those features.
3. Determine which models are the best.
4. Create a combined regression model, where we basically combine the features from the best models.
5. Use this final regression model to answer smart questions.
6. Report findings.

```{r}


cor(students[, c("Total_Score", "Study_Hours_per_Week", 
                         "Sleep_Hours_per_Night", 
                         "Attendance")], method="pearson")
```
```{r}
cor(students, method="pearson")
```

Target variables:
1. Pass_Fail (logistic regression)
2. Total_Score (linear regression?)

```{r}

```

```{r}
# Create individual regression models with features that seem to be correlated to the target variable
model_1 <- glm(Pass_Fail ~ Study_Hours_per_Week, 
                   data = students, family = binomial)

model_2 <- glm(Pass_Fail ~ Internet_Access_at_Home, 
                   data = students, family = binomial)

model_3 <- glm(Pass_Fail ~ Parent_Education_Level, 
                   data = students, family = binomial)

model_4 <- glm(Pass_Fail ~ Stress_Level , 
                   data = students, family = binomial)

model_5 <- glm(Pass_Fail ~ Extracurricular_Activities, 
                   data = students, family = binomial)

model_6 <- glm(Pass_Fail ~ Family_Income_Level, 
                   data = students, family = binomial)

model_7 <- glm(Pass_Fail ~ Study_Hours_per_Week * Family_Income_Level * Parent_Education_Level, 
                   data = students, family = binomial)
```

```{r}
head(predict(model_1, type = "response"))
head(predict(model_2, type = "response"))
head(predict(model_3, type = "response"))
head(predict(model_4, type = "response"))
head(predict(model_5, type = "response"))
head(predict(model_6, type = "response"))

head(predict(model_7, type = "response"))
```
```{r}
pred_probs_2 <- predict(model_2, type = "response")
pred_classes_2 <- ifelse(pred_probs_2 > 0.5, "Pass", "Fail")
pred_classes_2 <- factor(pred_classes_2, levels = c("Fail", "Pass"))

conf_mat_2 <- confusionMatrix(pred_classes_2, students$Pass_Fail, positive = "Pass")
cat("Model 2 - Internet Access\n")
cat("Precision:", round(conf_mat_2$byClass["Precision"], 3), "\n")
cat("Recall:", round(conf_mat_2$byClass["Recall"], 3), "\n\n")
```

```{r}
summary(model_1)
summary(model_2)
```


```{r}
model_lm <- lm(Total_Score ~ Study_Hours_per_Week + 
                        Sleep_Hours_per_Night + 
                        Attendance + 
                        Family_Income_Level + 
                        Parent_Education_Level + 
                        Extracurricular_Activities, 
                  data = students)
```


```{r}
summary(model_lm)
```


Note that no coefficients besides the intercept are statistically significant.

Target variable:
 1. Total score
 2. Grade
 3. Pass/Fail
 4. High pass, low pass, fail
 5. Honors, not honors


```{r} 
colnames(students) 
```

# FEATURE ENGINEERING
# Finding the best threshold value for the model
```{r}
roc_obj <- roc(students$Pass_Fail, students$predictions)
coords(roc_obj, "best", ret = "threshold")  
```

# Logistic regression Model with INVOLVEMENT_RATIO
```{r}
library(pROC)
students$Involvement_Ratio <- (students$Participation_Score + students$Attendance + students$Study_Hours_per_Week) / 3

# Model
Involvement_model <- glm(Pass_Fail ~ Involvement_Ratio, data = students, family = "binomial")

# Predictions
students$pred_Involvement <- predict(Involvement_model, type = "response")
students$pred_class_Involvement <- factor(ifelse(students$pred_Involvement > 0.66, "Pass", "Fail"), levels = c("Fail", "Pass"))

# confusion matrix
confusionMatrix(students$pred_class_Involvement, students$Pass_Fail, positive = "Pass")

# ROC Curve
roc(students$Pass_Fail,
    students$pred_Involvement,
    plot = TRUE,
    print.auc = TRUE)

```

# Logistic regression Model with STUDY HOURS PER WEEK
```{r}
students$Pass_Fail <- factor(students$Pass_Fail, levels = c("Fail", "Pass"))

# Model
study_hours_model <- glm(Pass_Fail ~ Study_Hours_per_Week,
                         data = students, family = binomial)

# Predictions
students$predictions_study_hours <- predict(study_hours_model, type = "response")
students$predicted_class_study_hours <- ifelse(students$predictions_study_hours > 0.66, "Pass", "Fail")
students$predicted_class_study_hours <- factor(students$predicted_class_study_hours, levels = c("Fail", "Pass"))

# confusion matrix
confusionMatrix(students$predicted_class_study_hours, students$Pass_Fail, positive = "Pass")

# ROC Curve 
roc(students$Pass_Fail,
    students$predictions_study_hours,
    plot = TRUE,
    print.auc = TRUE,
    col = "red",
    lwd = 2)

```

# Logistic regression Model with STRESS_MODEL
```{r}

# Model
stress_model <- glm(Pass_Fail ~ Stress_Level, data = students, family = "binomial")

# Predictions
students$pred_stress <- predict(stress_model, type = "response")
students$pred_class_stress <- factor(ifelse(students$pred_stress > 0.66, "Pass", "Fail"), levels = c("Fail", "Pass"))

# confusion matrix
confusionMatrix(students$pred_class_stress, students$Pass_Fail, positive = "Pass")

# ROC Curve
roc(students$Pass_Fail,
    students$pred_stress,
    plot = TRUE,
    print.auc = TRUE,
    col = "red")
```

# Logistic regression Model with ATTENDANCE_MODEL
```{r}
# Model
attendance_model <- glm(Pass_Fail ~ Attendance, data = students, family = "binomial")

# Predictions
students$pred_attendance <- predict(attendance_model, type = "response")
students$pred_class_attendance <- factor(ifelse(students$pred_attendance > 0.66, "Pass", "Fail"), levels = c("Fail", "Pass"))

# confusion matrix
confusionMatrix(students$pred_class_attendance, students$Pass_Fail, positive = "Pass")

# ROC Curve
roc(students$Pass_Fail, students$pred_attendance,
    plot = TRUE,
    print.auc = TRUE,
    col = "purple",
    lwd = 2,
    main = "ROC Curve - Attendance Model")

```



